Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-09-01 manu-draft.Rmd
results text updated with better handling of nondetects/secchi on bottom, fix #1

data(rswqdat)
data(rstrnpts)
data(rstrndat)
data(trnsum)
load(file = here('tables/wqsumtab.RData'))
load(file = here('tables/stktab.RData'))
load(file = here('tables/wqcmptab.RData'))
load(file = here('tables/savtab.RData'))
load(file = here('tables/mcrtab.RData'))

@@

A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates shows significant assocations between groups of variables (Figure \@ref(fig:allpca)).  The first principal component axis explained 31\% of the variation among all variables, whereas the second and third axes explained 33\% of the remaining variation.  For the water quality varialbes, the first principal component axis had positive loadings for secchi depth and dissolved oyxgen and negatative assocations for total nitrogen, chl-a, and temperature.  For macroalgae, the first axis was positively associated with frequency occurrence of red and green macroalgae and negatively associated with cyanobacteria. Loading vectors for the seagrasses were minor, although turtle grass (*T. testudinum*) had a weak positive association with axis 1.  The second principal component axis had a strong postive assocation with total phosphorus and negative association with salinity. The third axis showed a strong positive association with manatee grass (*S. filiforme*) and negative association with shoal grass (*H. wrightii*).

Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).  Chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH (all $p < 0.005$) and negatively correlated with secchi depth ($p < 0.005$).  Chlorophyll-a was also negatively correlated red and green macroalgae ($p < 0.05$) and turtle grass ($p < 0.005$).  Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas both were negatively correlated with cyanobacteria ($p < 0.005$ for red, $p < 0.05$ for green).  For water quality and seagrass, chl-a, total nitrogen, and total phosphorus was negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).  For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$ for both), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).  Among the seagrasses, manatee grass was negatively correlated with turtle grass and shoal grass ($p < 0.005$ for both).

@@

* Explain results - what's up with inorganic nutrients?  Likely driving phyto response but concentrations are not a good indicator.  Uptake is rapid...
* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress

>>>>

data(rswqdat)
data(rstrnpts)
data(rstrndat)
data(trnsum)
data(allpcadat)

load(file = here('tables/wqsumtab.RData'))
load(file = here('tables/stktab.RData'))
load(file = here('tables/wqcmptab.RData'))
load(file = here('tables/savtab.RData'))
load(file = here('tables/mcrtab.RData'))

@@

```{r}
expvr <- allpcadat$eig[, 2] %>% 
  round(0)
```

A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates showed significant associations between groups of variables (Figure \@ref(fig:allpca)).  The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]` of the remaining variation.  For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and strong negative loadings for total nitrogen, chl-a, and temperature.  For macroalgae, the first axis had positive loadings with frequency occurrence of red and green macroalgae. For seagrasses, turtle grass (*T. testudinum*) had a strong positive loading with axis 1.  The second principal component axis had strong positive loadings for secchi depth, total phosphorus, and ammonia and negative loadings for salinity and cyanobacteria macroalgae.  The third axis showed a strong positive loading with shoal grass (*H. wrightii*) and negative association with manatee grass (*S. filiforme*).  Observations by area showed that samples in area 3 were grouped by positive loadings on on the first PCA axis and negative loadings on the second PCA axis and were distinct from points in area 1.

Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).  Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all, except $p < 0.05$ for pH) and negatively correlated with secchi depth ($p < 0.05$). Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$) and negatively correlated with salinity and secchi ($p < 0.05$).  Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas both were negatively correlated with cyanobacteria ($p < 0.005$ for red, $p < 0.05$ for green).  Among the seagrasses, manatee grass was negatively correlated wtih shoal grass and turtle grass ($p < 0.005$).  Several notable correlations were also observed between water quality, macroalgae, and seagrass. Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$).   For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).  For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).

@@

* Explain results - what's up with inorganic nutrients?  Likely driving phyto response but concentrations are not a good indicator.  Uptake is rapid...
* Explain correlations/PCA - area 3 distinct from area 1 (high salinity/turtle grass dominated, likely just a natural grouping)
* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress
